Designing Types for R, Empirically

### 2020-05-14 paper/main.tex
intro

xxx WIP xxxx

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to propose changes to the
language, and for any changes to be accepted by the user community they
must clearly benefit the language without endangering backwards
compatibility. Our goal is thus to find a compromise between simplicity and
usefulness; the proposed type system should cover most common programming
idioms while remaining easy to use. In order to do this, we need to
understand the degree of polymorphism present in R code, that is to say, how
programmers leverage the dynamic nature of R to write code that can accept
arguments of different types.  This understanding will drive our design.

@@

How does one retrofit a type system to an existing language? Three succesful
models are Hack, Typed Racket and TypeScript.

The story that we're aiming for: Designing type systems for established
languages can be a challenging.  We aim to approach this problem by
essentially reverse-engineering a type system from a corpora analysis of a
representative subset of R code.  We present a simple type system that
captures the vast majority of usage of R --- it is lightweight, and
\AT{should provide enough information to be useful to a compiler}.

To evaluate our type system, we present a type-checker that is aware of R's
promise semantics.  In R, function arguments are wrapped in promises such
that they are not evaluated until they are needed.  Our tool deals with
these promises by inserting a check which fires only when the promise is
forced, at which point the value is checked for compliance with the type.

%
%
%
\section{Background}

@@

%
%
%
%
\subsection{The R Programming Language}
\label{sec:R}

@@

\begin{figure}[!hb]{\small
\begin{lstlisting}
> normSum <- function( m )  round( m / sum(m), 2)
> normSum(c(1L,3L,6L))
[1] 0.1 0.3 0.6
> normSum(c(1.1,3.3,6.6))
[1] 0.1 0.3 0.6
> normSum(c(1.6,3.3,6.1))
[1] 0.15 0.30 0.55
> normSum(complex(r=rnorm(3),i=rnorm(3)))
[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
\end{lstlisting}}
\caption{Sample R code}\label{sample}
\end{figure}

@@

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|c|l@{}}\hline
\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & \L & vector of boolean values\\
\k{integer}   & \I & vector of 32 bit integer values\\
\k{double}    & \D & vector of 64 bit floating points\\
\k{complex}   & \X & vector of complex values\\
\k{character} & \C & vector of strings values\\
\k{raw}       & \R & vector of bytes\\
\k{list}      & \l & vector of values of any type\\\hline
\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      & \sN &  singleton null value\\
\k{S4}        & \sS &  instance of a S4 class \\
\k{closure}   & \sF & a function with its environment\\
\k{environment}&\sE &  a mapping from symbol to value \\\hline
\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
\multicolumn{3}{l}{\k{special},
\k{builtin},
\k{symbol} (\sY),
\k{pairlist},
\k{promise}}\\
\multicolumn{3}{l}{
\k{language},
\k{char},
\k{...},
\k{any},
\k{expression},
}\\
\multicolumn{3}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

@@

\paragraph{Summary.} The most common values in R computations are vectorized
types. R programs do not have a way to constrain values to be scalar.
\k{NULL} is sometimes used to represent the case when no value is
available. \k{NA} is used within vector to represent missing observations.
Attributes can decorate values and are used as building blocks for
object-oriented programming. A potential type system for R could focus only
on the builtin types, if one wanted to strive for simplicity, or it could
try to capture attributes at the risk of increased complexity.

%
%
%
%
\subsection{R-dyntrace}
\label{sec:r-dyntrace}

R-dyntrace is an efficient dynamic analysis framework for R.

%
%
%
%
\section{Project Corpus}\label{sec:corpus}

For this paper we have selected \CorpusLoadable packages consisting of
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
code (C/C++/Fortran).  \FK{Add more details} Figure~\ref{fig:corpus} shows
all the packages, the size of the dots reflects the size of the project in
terms of both R and native lines of code\footnote{ Lines of source code
  reported excludes comments and blank lines, counted by \emph{cloc}, \cf
  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
expression code coverage in percents and the y-axis gives the number of
reverse dependencies in log scale. Dotted lines indicate means.  Packages
with over \PackageSizeOutierRnd lines of code are annotated.

@@

%
%
%
%
%
%
\section{Methodology}

@@

%
%
%
%
\subsection{typetracer and Initial Analysis}

@@

\item

First, we wanted to ensure that \typetracer captured types for unevaluated arguments.  
Unfortunately, not every argument of a function is necessarily used during every function execution, meaning that some arguments may remain unevaluated (as they are never forced), and there are a few reasons for this.

@@

\item Third, we had to contend with R's C implementation sometimes
  performing longjumps, which discards current R calls from the call stack.
  In order to ensure that call traces are not lost when a longjump occurs,
  \typetracer intercepts the unwinding process initiated by the longjump and
  mimics the functionality of the functions being exited.  The R-dyntrace
  framework supplies the return value during the unwinding process (if there
  is indeed such a value), so we can capture it, analyze it, and include it
  in the call trace.  In certain situations, however, no return value can be
  obtained (e.g. a longjump occurred as part of some error handling), and
  then we simply record a ``jumped'' return value, which we can deal with in
  post-processing.  
  % These longjumps occur quite often, particularly when S3 dispatch occurs, as the R implementation optimizes returns from dispatched functions by initiating a longjump.

>>>>

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to eventually propose changes to the
language, but we are aware that for any changes to be accepted by the user
community they must clearly benefit the language without endangering
backwards compatibility. Our goal is thus to find a compromise between
simplicity and usefulness; any proposed type system should cover common
programming idioms while remaining easy to learn and to use.

In order to do this, we need to understand the degree of polymorphism
present in R code, that is to say, how programmers leverage the dynamic
nature of R to write code that can accept arguments of different types.
This understanding will drive our design.

@@

%
\section{Background} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%
%
\subsection{The R Programming Language}\label{sec:R}

@@

\begin{figure}[!hb]{\begin{lstlisting}
> normSum <- function( m )  round( m / sum(m), 2)
> normSum(c(1L,3L,6L))
[1] 0.1 0.3 0.6
> normSum(c(1.1,3.3,6.6))
[1] 0.1 0.3 0.6
> normSum(c(1.6,3.3,6.1))
[1] 0.15 0.30 0.55
> normSum(complex(r=rnorm(3),i=rnorm(3)))
[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
\end{lstlisting}}
\caption{Sample R code}\label{sample}
\end{figure}

@@

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|l@{}}\hline
\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double}    & vector of 64 bit floating points\\
\k{complex}   & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw}       & vector of bytes\\
\k{list}      & vector of values of any type\\\hline
\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      &  singleton null value\\
\k{S4}        &  instance of a S4 class \\
\k{closure}   &  function with its environment\\
\k{environment}& mapping from symbol to value \\\hline
\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
\multicolumn{2}{l}{\k{special},
\k{builtin},
\k{symbol},
\k{pairlist},
\k{promise}}\\
\multicolumn{2}{l}{
\k{language},
\k{char},
\k{...},
\k{any},
\k{expression},
}\\
\multicolumn{2}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

@@

%
%
\section{Project Corpus}\label{sec:corpus}

For this paper we have selected \CorpusLoadable packages consisting of
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
code (C/C++/Fortran). Figure~\ref{fig:corpus} shows all the packages, the
size of the dots reflects the size of the project in terms of both R and
native lines of code\footnote{ Lines of source code reported excludes
  comments and blank lines, counted by \emph{cloc}, \cf
  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
expression code coverage in percents and the y-axis gives the number of
reverse dependencies in log scale. Dotted lines indicate means.  Packages
with over \PackageSizeOutierRnd lines of code are annotated.

@@

%
%
%
\section{Methodology}

@@

%
%
\subsection{typetracer and Initial Analysis}

@@

\item First, we wanted to ensure that \typetracer captured types for
  unevaluated arguments.  Unfortunately, not every argument of a function is
  necessarily used during every function execution, meaning that some
  arguments may remain unevaluated (as they are never forced), and there are
  a few reasons for this.

@@

\item Third, we had to contend with R's C implementation sometimes
  performing longjumps, which discards current R calls from the call stack.
  In order to ensure that call traces are not lost when a longjump occurs,
  \typetracer intercepts the unwinding process initiated by the longjump and
  mimics the functionality of the functions being exited.  The R-dyntrace
  framework supplies the return value during the unwinding process (if there
  is indeed such a value), so we can capture it, analyze it, and include it
  in the call trace.  In certain situations, however, no return value can be
  obtained (e.g. a longjump occurred as part of some error handling), and
  then we simply record a ``jumped'' return value, which we can deal with in
  post-processing.  
\end{itemize}
